A Bayesian approach to burial networks to infer social changes in 17th northeastern Taiwan during the European colonization period

### 2021-01-04 analysis/paper/paper.Rmd
source R scripts in the Rmd

<!--burial study- network analysis-->

Burials analysis by archaeologists is an approach to understand past social structure through the study of the physical traces of mortuary practices. The material cultures and biological records of burial behaviors can represent the social ranking or identity of the deceased [@Binford1971; @Drennan2010; @Saxe1970]. Despite criticism that manipulation of burial rituals by the living can cause a disconnect between a person's status in life and their status represented by burial contexts [@Hodder1980; @Pearson1982], burials can still provide valuable information to infer past societies [@Chapman2003]. For example, mortuary practices, including burial forms, grave goods, and ritual behaviors, that structure the material configuration of burials can represent social relations between members of a community, allowing the inference of social structures [@Byrd1995; @Seikel2011]. Based on network science and graph theories, the recent development of social network analysis provides many new tools to visualize and analyze relationships for archaeological data [@Borgatti2009; @Brughmans2013]. Network science is the study of relational data where a phenomenon is conceptualized as a network through two steps, abstraction and representation, based on our specification in terms of network theories [@Brandes2013; @Collar2015]. That is, a network can be viewed as a patterned aggregation that includes individual elements (i.e. individual burials), pair-wise relationships (the dyads, for example burials with similar types and amounts of grave goods), and an overall structure showing global patterns represented in the data.

@@

```{r read-data}
library(readxl)
library(tidyverse)
library(here)

# read data
burial <- read_excel(here("analysis", "data", "raw_data", "Kiwulan_Burials.xlsx"))
```

```{r tidy-data}
# number of each phase
burial_phases_sample_size <-
  burial %>%
  mutate(Phase = ifelse(Phase == 'euro', 'post', Phase)) %>%
  filter(!is.na(Phase)) %>% 
  count(Phase)

pre_E <- burial_phases_sample_size %>% filter(Phase == "pre") %>% pull()
post_E <- burial_phases_sample_size %>% filter(Phase == "post") %>% pull()
chi <- burial_phases_sample_size %>% filter(Phase == "chi") %>% pull()
diff_pre_post <- post_E - pre_E
```

@@

@@

<!-- burial data and the definition of tie formation  --> 
We analyzed burial data collected from the published excavation reports, and the original fieldwork notes of Kiwulan. Trade beads are commonly found across burials with substantial differences in quantities, so we described each burial as having one of four levels, high, upper-middle, lower-middle, and low, according to their distributions across all burials. Gold-foil beads are in levels of high (>3), upper-middle (=3), lower-middle (=2), and low (=1); carnelian beads are in levels of high (>6), upper-middle (=4-6), lower-middle (=3), and low (=1-2); glass beads are in levels of high (>6), upper-middle (=3-6), lower-middle (=2), and low (=1). That means if burial 1 and burial 2 both have high quantities of agate beads, then there will be a tie connecting them. For less frequent prestige goods, including imported porcelains, gold foils, and fish-shaped ornaments, we linked two burials when they possess each type of goods in common (i.e. presence or absence). Node attributes here include osteological data, such as age and sex, and cultural data, such as ritual pottery, and burial value index. Ritual pottery was identified as locally made ceramics placed above graves that suggests funeral feasting according to historical records. The burial value index is an important attribute for economical inequality that we assigned the burials into three classes, high (>30), medium (>11 and <30), and low (<12), as an index of wealth. The number is the sum of values of each type of prestige good, which is calculate by the total number of a prestige item from burial context over the number of prestige item in a burial. Since burials tend to have multiple prestige goods in common, the network ties are weighted instead of binary data (where the value 1 represents a tie and the value 0 otherwise) [@Snijders2011]. For example, if two burials have both low quantity of glass beads and porcelain in common, the tie is given a value of 2. Our networks are non-directed, which means ties have no orientation forming the relationship between actors to indicate a mutual relationship.

@@

```{r bergm-iteration}
burn_in <- 200
main_iter <- 1000
aux_iter <- 10000
```

@@

<!-- summary of the research by answering questions raised in Intro, and anthropological contribution -->
In this paper, we presented a naval approach for studying burial relations using ERGM network analysis within a Bayesian framework to rigorously examined social changes in a European percicolonial context in northeastern Taiwan. We tested the hypotheses of chronological changes in interburial networks at Kiwulan with the evaluation of both endogenous and exogenous network effects. The results support our hypothesize model that the relationship between burials changed after European colonization period in the 17th century. Before the arrival of the Europeans, the network properties has a tendency of more closed burial relationships with ritual elements as the key formation mechanism. After the European arrival, the network has a tendency of centralization relative to the rarity of goods. The changes in formation mechanisms for networks from ritual-oriented practices to wealth-oriented possession suggest the behavior of wealth accumulation and an increased differentiation stimulated by the European presence and associated long-distance trade network. This aliens with changes in social complexity from a more corporate to a more network mode society that stress different strategies for achieving power and different degrees of social inequality, indicative of an increased social inequality. Using burial data with historical documents, we are able to detect the processes of changes in Indigenous societies in the periphery of European colonial bases to better understand indirect impacts that are underestimated in this region [@Acabado2017; @Trabert2017]. This study not only discuss pericolonal impacts that are not well understood yet, but also points out the agency of local people to manipulate colonial power instead of being passive recipients of trade goods.

@@

>>>>

<!--burial study- network analysis-->
Burials analysis by archaeologists is an approach to understand past social structure through the study of the physical traces of mortuary practices. The material cultures and biological records of burial behaviors can represent the social ranking or identity of the deceased [@Binford1971; @Drennan2010; @Saxe1970]. Despite criticism that manipulation of burial rituals by the living can cause a disconnect between a person's status in life and their status represented by burial contexts [@Hodder1980; @Pearson1982], burials can still provide valuable information to infer past societies [@Chapman2003]. For example, mortuary practices, including burial forms, grave goods, and ritual behaviors, that structure the material configuration of burials can represent social relations between members of a community, allowing the inference of social structures [@Byrd1995; @Seikel2011]. Based on network science and graph theories, the recent development of social network analysis provides many new tools to visualize and analyze relationships for archaeological data [@Borgatti2009; @Brughmans2013]. Network science is the study of relational data where a phenomenon is conceptualized as a network through two steps, abstraction and representation, based on our specification in terms of network theories [@Brandes2013; @Collar2015]. That is, a network can be viewed as a patterned aggregation that includes individual elements (i.e. individual burials), pair-wise relationships (the dyads, for example burials with similar types and amounts of grave goods), and an overall structure showing global patterns represented in the data.

@@

```{r import-and-inspect-data, include=FALSE, cache=TRUE, cache.extra=tools::md5sum(here::here("000-prep1-value-burials.R"))}
suppressWarnings(source(here::here("analysis", "paper", "000-prep1-value-burials.R")))
```

```{r tidy-data, include=FALSE, cache=TRUE, cache.extra=tools::md5sum(here::here("001-data-tidy.R"))}
suppressWarnings(source(here::here("analysis", "paper", "001-data-tidy.R")))
```

@@

```{r sample-size}
# number of each phase
library(tidyverse)
pre_E <- burial_three_period_age_number %>% filter(Phase == "pre") %>% pull()
post_E <- burial_three_period_age_number %>% filter(Phase == "post") %>% pull()
chi <- burial_three_period_age_number %>% filter(Phase == "chi") %>% pull()
diff_pre_post <- post_E - pre_E
```

@@

<!-- burial data and the definition of tie formation  --> 
We analyzed burial data collected from the published excavation reports, and the original fieldwork notes of Kiwulan. Trade beads are commonly found across burials with substantial differences in quantities, so we described each burial as having one of four levels, high, upper-middle, lower-middle, and low, according to their distributions across all burials. Gold-foil beads are in levels of high (> 3), upper-middle (= 3), lower-middle (= 2), and low (= 1); carnelian beads are in levels of high (> 6), upper-middle (= 4-6), lower-middle (= 3), and low (= 1-2); glass beads are in levels of high (> 6), upper-middle (= 3-6), lower-middle (= 2), and low (= 1). That means if burial 1 and burial 2 both have high quantities of agate beads, then there will be a tie connecting them. For less frequent prestige goods, including imported porcelains, gold foils, and fish-shaped ornaments, we linked two burials when they possess each type of goods in common (i.e. presence or absence). Node attributes here include osteological data, such as age and sex, and cultural data, such as ritual pottery, and burial value index. Ritual pottery was identified as locally made ceramics placed above graves that suggests funeral feasting according to historical records. The burial value index is an important attribute for economical inequality that we assigned the burials into three levels, high (>= 29, account for 10%), above average (= 12-28), and below average (< 12), as an index of wealth. The number is the sum of values of each type of prestige good, which is calculate by the total number of a prestige item from burial context over the number of prestige item in a burial. Since burials tend to have multiple prestige goods in common, the network ties are weighted instead of binary data (where the value 1 represents a tie and the value 0 otherwise) [@Snijders2011]. For example, if two burials have both low quantity of glass beads and porcelain in common, the tie is given a value of 2. Our networks are non-directed, which means ties have no orientation forming the relationship between actors to indicate a mutual relationship.

@@

```{r bergm-iteration}
burn_in <- 100
main_iter <- 1000
aux_iter <- 4000
```

@@

<!-- summary of the research by answering questions raised in Intro, and anthropological contribution -->
In this paper, we presented a naval approach for studying burial relations using ERGM network analysis within a Bayesian framework to rigorously examined social changes in a European pericolonial context in northeastern Taiwan. We tested the hypotheses of chronological changes in interburial networks at Kiwulan with the evaluation of both endogenous and exogenous network effects. The results support our hypothesize model that the relationship between burials changed after European colonization period in the 17th century. Before the arrival of the Europeans, the network properties has a tendency of more closed burial relationships with ritual elements as the key formation mechanism. After the European arrival, the network has a tendency of centralization relative to the rarity of goods. The changes in formation mechanisms for networks from ritual-oriented practices to wealth-oriented possession suggest the behavior of wealth accumulation and an increased differentiation stimulated by the European presence and associated long-distance trade network. This aliens with changes in social complexity from a more corporate to a more network mode society that stress different strategies for achieving power and different degrees of social inequality, indicative of an increased social inequality. Using burial data with historical documents, we are able to detect the processes of changes in Indigenous societies in the periphery of European colonial bases to better understand indirect impacts that are underestimated in this region [@Acabado2017; @Trabert2017]. This study not only discuss pericolonal impacts that are not well understood yet, but also points out the agency of local people to manipulate colonial power instead of being passive recipients of trade goods.

@@

We would like to thank the Yilan County Cultural Affairs Bureau in Taiwan for permitting access to the data used in this paper. We thank the excavation team of Kiwulan led by Yu-pei Chen for their post-excavation analysis and detailed documents about burials that enables this study. We also thank Alberto Caimo and Steven M. Goodreau for the discussion regarding the methods at the earliest stage fo analysis. This work was supported in part by the travel grant from the Department of Anthropology at University of Washington and the Doctoral Dissertation Fellowship from the Chiang Ching-kuo Foundation for International Scholarly Exchange (Project #DF009-A-18). We thank Ben Fitzhugh and Peter Lape for their insightful comments on an early draft. We also acknowledge comments from anonymous reviewers that have greatly improved this manuscript.
